msprime 1.0: an efficient and extensible coalescent simulation framework

### 2020-09-27 paper.tex

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
\usepackage{showkeys}

@@

%% local definitions for section multiple merger coalescents
\newcommand{\msprime}[0]{{\texttt{msprime} }}
\newcommand{\tskit}[0]{{\texttt{tskit} }}
\newcommand{\ms}[0]{{\texttt{ms} }}
\newcommand{\scrm}[0]{{\texttt{scrm} }}
\newcommand{\stdpopsim}[0]{{\texttt{stdpopsim} }}
 \newcommand{\be}{\begin{equation}}
 \newcommand{\ee}{\end{equation}}
 \newcommand{\bd}{\begin{displaymath}}
 \newcommand{\ed}{\end{displaymath}}
\newcommand{\IN}{\ensuremath{\mathds{N}}}%
\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%

@@

\subsection*{Ancestral recombination graphs}
The definition of ARGs is vague and confusing.
\cite{minichiello2006mapping} define it as the data structure
rather than the original stochastic
process~\citep{griffiths1991two,griffiths1997ancestral}.
\cite{rasmussen2014genome} have a slightly different approach.
Sometimes the ARG is useful.
The tree sequence output for \msprime\ can be turned into
an ARG using the \texttt{record\_full\_arg} option, where we store
recombination nodes.

We also provide an interface to compute the likelihood of a given ARG
under the~\cite{kuhner2000maximum} model.

\subsection*{Instantaneous bottlenecks}
A common approach to modelling the effect of demographic history on genealogies is to assume that effective population size ($N_e$) changes in discrete steps which define a series of epochs~\citep{griffiths1994sampling, marth2004allele, keightley2007joint, li2011inference}. In this setting of piece-wise constant $N_e$, capturing a population bottleneck requires three epochs: $N_e$ is reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If bottlenecks are short both on the timescale of coalescence and mutations, one may expect little information about the duration of a bottleneck (T_end-T_end) in sequence data. Thus a simpler, alternative model is to assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$) and generate a sudden burst of coalescence events (a multiple merger event) in the genealogy. The strength of the bottleneck B can be thought of as an (imaginary) time period during which coalescence events are collapsed, i.e. there is no growth in genealogical branches during B and the probability that a single pair of lineages entering the bottleneck coalesce during the bottleneck is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is attractive and both analytic results and empirical inference~\citep{griffiths1994sampling, galtier2000detecting, bunnefeld2015inferring} have been developed under this model, it has not been straightforward to simulate data under instantaneous bottleneck histories with \ms\. Instantaneous bottlenecks are implemented as a demographic event in \msprime\.

@@

\section*{Acknowledgments}
JK is supported by the Robertson Foundation. [Others fill in their ACKs please]

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{listings}
\usepackage[english]{babel}%
\usepackage[T1]{fontenc}%
\usepackage[utf8]{inputenc}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{geometry}%
\usepackage{dsfont}
\usepackage{verbatim}%
\usepackage{environ}%
\usepackage[right]{lineno}%
%\usepackage{showkeys}

@@

%% local definitions for section multiple merger coalescents
 \newcommand{\be}{\begin{equation}}
 \newcommand{\ee}{\end{equation}}
 \newcommand{\bd}{\begin{displaymath}}
 \newcommand{\ed}{\end{displaymath}}
\newcommand{\IN}{\ensuremath{\mathds{N}}}%
\newcommand{\EE}[1]{\ensuremath{\mathds{E}\left[ #1 \right]}}%
\newcommand{\one}[1]{\ensuremath{\mathds{1}_{\left\{ #1 \right\}}}}%
\newcommand{\prb}[1]{\ensuremath{\mathds{P}\left( #1 \right) } }%

@@

\subsection*{Ancestral recombination graphs}

The \msprime~simulator generates tree sequences by simulating realisations of the
Ancestral Recombination Graph (ARG).
By default, only the marginal genealogy at each site is recorded in the tree sequence
output.
Marginal genealogies are sufficient for e.g.~simulating and storing sequence data,
and involve fewer edges than nodes than storing a realisation of the full ARG.
However, the mapping from ARG realisations to marginal tree sequences is not lossless.
For example, times of recombination events can be recovered from the former
but not the latter.
The same is true of the number of extant ancestors at each time in the past.
In addition to the default behaviour, we support the retention of the full
ARG realisation using the \texttt{record\_full\_arg} option.
This option augments the default tree sequence data structure with three
new kinds of nodes:
\begin{itemize}
\item Each recombination event is marked by two nodes with identical birth times
given by the time of the recombination event, and marked with the
\texttt{NODE\_IS\_RE\_EVENT} flag. 
Each node represents an ancestor of the recombinant child node, and edges connect
the relevant segments of genome from the child node to these two ancestors.
\item We store common ancestor nodes even when the corresponding merger event
does not result in coalescence of ancestral material.
The resulting nodes are marked with the \texttt{NODE\_IS\_CA\_EVENT} flag.
\item When multiple populations are simulated, migration events are depicted by nodes
with the \texttt{NODE\_IS\_MIG\_EVENT} flag, and with birth times equal to the time of
the migration.
The population flag of the migration node identifies the population to which the lineage
belongs after the reverse time migration event.
\end{itemize}

The likelihood, i.e.~the joint sampling probability of the ancestry and mutations,
is an example of a function which is intractable when the ancestry is stored as a
marginal tree sequence, but tractable when the ancestry is given by a full
ARG realisation.
The likelihood is a central quantity in evaluating the plausibility of a putative
ancestry as an explanation of DNA sequence data, both directly through
e.g.~approaches based on maximum likelihood, and as an ingredient of
methods such as Metropolis-Hastings
\citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian}.

We provide two functions to facilitate likelihood-based inference.
Both are implemented only for the simplest case of the standard ARG with a 
constant population size, and require tree sequences compatible with the 
\texttt{record\_full\_arg} option as their arguments.
\begin{itemize}
\item \texttt{msprime.log\_arg\_likelihood(ts, r, N)} returns the natural logarithm of 
the sampling probability of the tree sequence \texttt{ts} under the ARG with per-link,
per-generation recombination probability \texttt{r} and population size \texttt{N}.
Specifically, the function returns the logarithm of
\begin{equation*}
\Bigg( \frac{ 1 }{ 2 N } \Bigg)^{ q_c } \Bigg( \prod_{ i : \mathcal{R} } r g_i \Bigg)
	\exp\Bigg( -\sum_{ i = 1 }^q \Big[\frac{ 1 }{ 2 N } \binom{ k_i }{ 2 } 
		+ r l_i \Big] t_i  \Bigg),
\end{equation*}
where $t_i$ is the number of generations between the $(i - 1)$th and $i$th event,
$k_i$ is the number of extant ancestors in that interval, $l_i$ is the number of links
in that interval that would split ancestral material should they recombine,
$q$ is the total number of events in the tree sequence $\texttt{ts}$,
$q_c$ is the number of coalescences, $\mathcal{R}$ is the set of indices 
of time intervals which end in a recombination, and $g_i$ is the corresponding
\emph{gap}:  the length of material in which the recombination at the end of the 
$i$th period could have taken place without affecting the resulting genealogy.
A recombination in ancestral material will always have $g_i = 1$, while a recombination
in trapped non-ancestral material can result in other values of $g_i$.
For a continuous model of the genome and a recombination in ancestral material,
we set $g_i = 1$ and interpret the result as a density.
\item \texttt{msprime.unnormalised\_log\_mutation\_likelihood(ts, m)} returns the
natural logarithm of the probability of the mutations recorded in the tree sequence
\texttt{ts} given the corresponding ancestry, assuming the infinite sites model, up to
a normalising constant which depends on the pattern of mutations,
but not on the tree sequence or the per-site, per-generation mutation 
probability \texttt{m}.
Specifically, the function returns the logarithm of
\begin{equation*}
e^{ - T m / 2 } \frac{ ( T m / 2 )^M }{ M ! }
\prod_{ \gamma \in \mathcal{ M } } \frac{ h_{ \gamma } }{ T },
\end{equation*}
where $T$ and $\mathcal{M}$ are the total branch length and set of mutations
in \texttt{ts}, respectively, and for a mutation $\gamma$, $h_{ \gamma }$ is the
total branch length on which $\gamma$ could have arisen while appearing on all
of the leaves of \texttt{ts} it does, and on no others.
Unary nodes on marginal trees arising from the \texttt{record\_full\_arg} option
mean that, in general $h_{ \gamma }$ corresponds to the length of one or more
edges.
\end{itemize}

\begin{comment}
Previous draft of the ARG feature retained here as a comment.

The definition of ARGs is vague and confusing.
\cite{minichiello2006mapping} define it as the data structure
rather than the original stochastic
process~\citep{griffiths1991two,griffiths1997ancestral}.
\cite{rasmussen2014genome} have a slightly different approach.
Sometimes the ARG is useful.
The tree sequence output for \msprime\ can be turned into
an ARG using the \texttt{record\_full\_arg} option, where we store
recombination nodes.

We also provide an interface to compute the likelihood of a given ARG
under the~\cite{kuhner2000maximum} model.
\end{comment}

\subsection*{Instantaneous bottlenecks}
A common approach to modelling the effect of demographic history on genealogies is to assume that effective population size ($N_e$) changes in discrete steps which define a series of epochs~\citep{griffiths1994sampling, marth2004allele, keightley2007joint, li2011inference}. In this setting of piece-wise constant $N_e$, capturing a population bottleneck requires three epochs: $N_e$ is reduced by some fraction $b$ at the start of the bottleneck, $T_{start}$, and recovers to its initial value at time $T_{end}$~\citep{marth2004allele}. If bottlenecks are short both on the timescale of coalescence and mutations, one may expect little information about the duration of a bottleneck ($T_{end}-T_{start}$) in sequence data. Thus a simpler, alternative model is to assume that bottlenecks are instantaneous ($T_{end}-T_{start} \rightarrow 0$) and generate a sudden burst of coalescence events (a multiple merger event) in the genealogy. The strength of the bottleneck B can be thought of as an (imaginary) time period during which coalescence events are collapsed, i.e. there is no growth in genealogical branches during B and the probability that a single pair of lineages entering the bottleneck coalesce during the bottleneck is $1-e^{-B}$. Although this simple two parameter model of bottlenecks is attractive and both analytic results and empirical inference~\citep{griffiths1994sampling, galtier2000detecting, bunnefeld2015inferring} have been developed under this model, it has not been straightforward to simulate data under instantaneous bottleneck histories with \ms. Instantaneous bottlenecks are implemented as a demographic event in \msprime.

@@

\section*{Acknowledgments}
JK is supported by the Robertson Foundation. 
Jere Koskela is supported in part by EPSRC grant EP/R044732/1.
[Others fill in their ACKs please]
